---
output: github_document
---

# From static to animated timeseries: the R way

In the last months, many examples on how to plot timeseries of COVID-19 confirmed cases can be found all over the web and you might be overloaded by the amount of data visualisations that each day are shared on every platform.
It is indeed hard, for any passionate in data visualisation to not want to plot the data yourself.
For this reason, in this article, we will go through the necessary steps to create a timeseries plot, inspired from the many examples that can be found on the web: FT, Economist.
As a second step we will then animate the chart, always in R, using the package `{gganimate}`.

## Importing the data and the libraries

As a very first step in our R script or R notebook, we will want to import all the libraries that we use in this little excerise. We start by the omnipresent `{tidyverse}`, which we will use to import, manipulate and plot the data.
`{lubridate}` is our package of choice for working with variables that store dates values, `{janitor}` is used to clean the column names. `{rcartocolor}` contains some very nice color scales, useful for cartography and not only, then we use `{gganimate}` to finally animate the chart.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
library(gganimate)
library(rcartocolor)
```

We can now import the data directly from the continuously updated dataset put together by the John Hopkins University, shared on Github.

```{r message=FALSE, warning=FALSE}
cases_raw <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv")
```


In order to recreate the chart, inspired by John Burn-Murdoch work, we need to tidy the data first, and then do some basic manipulation on it. 
Since each column corresponds to the different dates, the first thing to do would be to transform the data from wide format to long format: for this we use the `pivot_longer()` function from the `{tidyr}` package. 
We even tried to be smart here, and select for our pivoting, only the columns that are named as a date in the following format "digit/digit/digit".
The new column `date` needs to be transformed into a date column, and for doing it we can use the smart `mdy()` function from the `{lubridate}` package: where `m` stands for *month*, `d` stands for `day` and `y` stands for `year`.

In the rest of the script we keep only those `country_regions` that we want to plot. We also filter out the China data because it is a bit off scale compared to the other countries, but this may not be valid in few weeks time given the spread of the pandemic.

Since we have that some countries are ungrouped by subregions, we also use the `group_by()` and `summarise()` to make sure that we have only one value per day per country, which corresponds to the sum of cases of all the regions in each country.


```{r}
cases_data <- cases_raw %>%
  pivot_longer(cols= matches("\\d+/\\d+/\\d+"), 
               names_to = "date", 
               values_to = "cases") %>% 
  clean_names() %>% 
  transmute(country_region, date = mdy(date), cases) %>% 
  filter(!country_region %in% c("Others", "China", "Cruise Ship")) %>%
  group_by(country_region, date) %>% 
  summarise(cases = sum(cases))
```

If we just wanted to plot the raw timeseries, this would be enough. We although want to go a bit further than this and plot the dates since the 100th case in each country.
We also want to keep only the countries that reached the 100th case from less than 6 days.
Still by group, we can calculate the days since the 100th case and after this we can `ungroup` the data.

```{r}
cases_data <- cases_data %>% 
  filter(cases >= 100)%>%
  filter(n() >= 5) %>%
  arrange(date, .by_group = TRUE)%>%
  mutate(days_since_100 = (date-first(date)) / ddays(1)) %>%
  ungroup()
```

Many visualisations that can be found on the web or in newspapers, overlap to the country trends a dashed line that corresponds to the 33% daily rise. We then add to our original `tibble` (which is a table in `tidyverse` language).  

```{r}
cases_data <- cases_data %>% 
  bind_rows(
    tibble(country_region = "33% daily rise", 
           days_since_100 = 0:25) %>%
      mutate(cases = 100*1.33^days_since_100)
  ) %>% 
  mutate(line_type = ifelse(country_region == "33% daily rise", "2", "1")) 
```

We also create a new variable `line_type` that we will use to set the line corresponding to the `33% daily rise` to be a dashed line.


```{r}
ggplot(data = cases_data, 
       mapping = aes(x = days_since_100, 
                     y = cases, 
                     color = country_region)) +
  #geom_point(size = 0.8, alpha = 0.9, pch = 21)+
  geom_line(size = 0.7, alpha =0.9,
            mapping = aes(linetype = line_type)) +
  scale_y_log10(expand = expansion(add = c(0,0.1)), 
                breaks=c(100, 500, 2000,  10000, 60000)) +
  scale_color_carto_d(palette = "Safe")+
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "none",
    plot.margin = margin(3,15,3,3,"mm")
  ) +
  coord_cartesian(clip = "off") +
  labs(x = "Number of days since 100th case", 
       y = "Total Number of Cases",
       title = "Total number of COVID-19 cases",
       subtitle =  "Outside of China",
       caption = "Data Source: John Hopkins University")
```

This was a first attempt but we can already see that there is one problem with this visualisation: the colours. The number of countries plotted is indeed too big for many categorical colour scale. If we plotted each line in a different colour we would not be able to really discern the countries. If we want to still use different colours for the countries that we want to highlight we need to give an order to the countries, based on the data (i.e. based on the maximum value of `days_since_100`, for each country).
Do do so we can use the `fct_infreq()` from the `{forcats}` package.



```{r}
cases_data <- cases_data %>% 
  mutate(country_region = fct_infreq(country_region))
```


```{r}
ggplot(data = cases_data, 
       mapping = aes(x = days_since_100, 
                     y = cases, 
                     color = country_region)) +
  #geom_point(size = 0.8, alpha = 0.9, pch = 21)+
  geom_line(size = 0.7, alpha =0.9,
            mapping = aes(linetype = line_type)) +
  scale_y_log10(expand = expansion(add = c(0,0.1)), 
                breaks=c(100, 500, 2000,  10000, 60000)) +
  scale_color_carto_d(palette = "Safe")+
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "none",
    plot.margin = margin(3,15,3,3,"mm")
  ) +
  coord_cartesian(clip = "off") +
  labs(x = "Number of days since 100th case", 
       y = "Total Number of Cases",
       title = "Total number of COVID-19 cases",
       subtitle =  "Outside of China",
       caption = "Data Source: John Hopkins University")
```

## Annotating the plot

If we want to annotate the plot with the name of each country, we can then use `geom_text()` in our `ggplot` call. We can place the annotation next to the last point of each line, by using a different `tibble` for the annotation.
In `annotations` we will just keep the observation corresponding to the maximum value of `days_since_100`, for each county.
If we want to only annotate the first 12 countries (12 because the colour scale we have chosen has 12 colours), we can create a new variable named `label_country`, that will be empty for all countries apart from the first 12, ordered  before using the `fct_reorder()`.

```{r}
annotations <- cases_data %>% 
  group_by(country_region) %>% 
  filter(days_since_100 == max(days_since_100)) %>% 
  mutate(label_country = ifelse(country_region %in% levels(country_region)[1:12],
                                 as.character(country_region), ""))
```

Now we are ready to annotate the plot. To increase the visibility of the labels we can use `geom_shadowtext()` from the `{shadowtext}` package, which adds a little 3d subtle look, which makes the labels stick out a little bit.
We indeed all know, that the devil lies in the details.


```{r}
ggplot(data = cases_data, 
       mapping = aes(x = days_since_100, 
                     y = cases, 
                     color = country_region)) +
  geom_line(size = 0.7, alpha =0.9,
            mapping = aes(linetype = line_type)) +
  scale_y_log10(expand = expansion(add = c(0,0.1)), 
                breaks=c(100, 500, 2000,  10000, 60000)) +
  shadowtext::geom_shadowtext(data = annotations, 
                              mapping = aes(x = days_since_100, 
                                            y = cases, 
                                            label = label_country),
                              hjust=-0.1, vjust = 0, bg.color = "white") +
  scale_color_carto_d(palette = "Safe")+
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "none",
    plot.margin = margin(3,15,3,3,"mm")
  ) +
  coord_cartesian(clip = "off") +
  labs(x = "Number of days since 100th case", 
       y = "Total Number of Cases",
       title = "Total number of COVID-19 cases",
       subtitle =  "Outside of China",
       caption = "Data Source: John Hopkins University")
```

Now, how do we plot the coloured line on top of the grey lines?
To do this, we would need to reorder the `country_region` factors again, but this would affect the way the the lines are coloured.
How can we cope with this?

The way to do it is to manually set the colour of the countries, and then order the countries in the reverse order.


```{r message=FALSE, warning=FALSE}
cases_data <- cases_data %>% 
  mutate(color_label = forcats::fct_collapse(country_region,
    "#D63D32" = "Switzerland",
    "#888888" = "Italy",
    "#6699CC" = "Iran",
    "#661100" = "Germany",
    "#882255" = "France",
    "#999933" = "United Kingdom",
    "#44AA99" = "US",
    "#332288" = "Spain",
    "#117733" = "Korea, South",
    "#DDCC77" = "Netherlands",
    "#CC503E" = "Hong Kong",
    "#1D6996" = "Singapore",
    "#855C75" = "Japan",
    "black"   = "33% daily rise",
    other_level = "grey90"),
    color_label = fct_relevel(color_label, "grey90")) %>%
  arrange(color_label) %>%
  mutate(country_region = fct_inorder(country_region)) %>% 
  mutate(country_label = ifelse(color_label == "grey90", "", as.character(country_region)))
```

 


```{r}
ggplot(data = cases_data, 
       mapping = aes(x = days_since_100, 
                     y = cases, 
                     color = color_label,
                     group = country_region)) +
  geom_line(mapping = aes(linetype = line_type), 
            size = 0.7, alpha =0.9) +
  scale_y_log10(expand = expansion(add = c(0,0.1)), 
                breaks=c(100, 500, 2000,  10000, 60000)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "none",
    plot.margin = margin(3,15,3,3,"mm")
  )+
  coord_cartesian(clip = "off") +
  labs(x = "Number of days since 100th case", 
       y = "Total Number of Cases",
       title = "Total number of COVID-19 cases",
       subtitle =  "Outside of China",
       caption = "Data Source: John Hopkins University")
```


```{r}
plot_gg +
  transition_reveal(days_since_100)
```

```{r}
#anim_save("covid.gif")
```

