---
title: "Making a map of COVID-19 incidence in Switzerland, with {ggplot2}"
output: github_document
---


In this article, we are going to visualise the spatial distribution of  COVID-19 incidence, in the last 14 days in Switzerland, by creating a thematic map, also known as *choropleth* map.
In the last article, we have explored how to visualise COVID-19 time series data, and we have animated the visualisation using `{gganimate}`.

This time, we are going to rely on `{sf}` and `{ggplot2}` as our main tools.

`{sf}`, which stands for simple feature, is the library that we are using to deal with spatial vectorial data, which is data that describes geographical geometries as a series of points, whose coordinates are the *longitude* and the *latitude*.
This may sound abstract...
This package is the go-to tool for importing, manipulating and plotting this kind of data.

We are using `{readxl}` to import the excel file, downloaded from the [Swiss Federal Office of Public Health Website](https://www.bag.admin.ch/bag/en/home/krankheiten/ausbrueche-epidemien-pandemien/aktuelle-ausbrueche-epidemien/novel-cov/situation-schweiz-und-international.html).
`{rcartocolor}` is the R library that includes nice looking colour scales, that have been developed for cartography.
As David Letterman would say, `{tidyverse}` needs no introduction. 

```{r  message = FALSE, warning = FALSE}
library(tidyverse)
library(sf)
library(rcartocolor)
library(readxl)
```


Let's start loading the data, using `read_excel()`, in which we can define the exact name of the `sheet` that we want to load, how many line we can `skip` and how many lines we want to `keep`.
We have one row per canton and a row for the title, which means that we need to keep only 27 rows (the Swiss cantons are 26).

We also clean a bit the column names using the `clean_names` from the `{janitor}` package and using `transmute` to rename the wanted columns and drop the others.

```{r}
covid_incidence <- read_excel("resources/200325_dati di base_grafica_COVID-19-rapporto.xlsx", 
    sheet = "COVID19 casi per cantone", skip = 6, 
    n_max = 27) %>% 
  janitor::clean_names() %>% 
  transmute(canton = cantone, incidence = incidenza_100_000_6)

head(covid_incidence)
```

We have now one variable which contains the canton codes, and one variable that contains the incidence per 100'000, by canton, of COVID-19 in the last 14 days.

We are now ready to load the shapefiles.

> Wait, what are the shapefiles?

Shapefiles, are the files that contain the shapes that we want to plot. We want to plot the cantonal data, so we need to get the swiss cantons shapes, which can be downloaded from
[here](https://www.bfs.admin.ch/bfs/en/home/services/geostat/swiss-federal-statistics-geodata/administrative-boundaries/generalized-boundaries-local-regional-authorities.html).

Shapefiles are actually a set of files, which contain different geographic infos (e.g. info on the projections). One of these files, has the `.shp` extension and this is the one that we are going to load.

> Beware that you need to have all the other files in the same folder.

Now we can therefore load 3 shapefiles, one that contains the swiss border shape, one that containts the shape of the major lakes of Switzerland and one that contains the swiss cantons.

Let's load them and see how they look like.


```{r}

swiss_border <- st_read("resources/g2l15.shp")

swiss_lakes <- st_read("resources/g2s15.shp")

swiss_cantons <- st_read("resources/G1K09.shp")
```


```{r}
class(swiss_cantons)
```

`swiss_cantons`, like the other data that we have just loaded, are stored in an `sf data.frame`, so we can manipulate it just like we manipulate tibbles (or data.frames) and it stores the geometries in a very tidy way: as a nested variable usually called `geometry`.
This will be your only *special* variable, the others (that are called attributes) will just be normal variables.

```{r}
head(swiss_cantons)
```

> The advantage of using `{sf}` as our main tool to deal with these data types? We can use `{ggplot2}` to plot it!


```{r}
ggplot()+
  geom_sf(data = swiss_cantons)
```


And, just like with any `{ggplot2}` chart, we can reason layer by layer when building it.
Let's now add the Swiss lakes on top of the canton shapes and use `theme_void()` to remove the background and the axis.
In this step, we can also make the cantons transparent by setting the `fill` argument to NA and add a light teal colour fill the lakes.


```{r}
ggplot()+
  geom_sf(data = swiss_cantons, fill = NA) +
  geom_sf(data = swiss_lakes,  fill = "#d1eeea", color = "#d1eeea") +
  theme_void()
```


Now, how to colour each canton by the shape of COVID-19 incidence?

We just need to join the `covid_incidence` tibble and `swiss_cantons` table together, using the canton code as joining variable.
This will allow us to map the variable incidence to the `fill` aesthetic and create a **choropleth** map, i.e. a thematic map.

```{r}
swiss_cantons <- swiss_cantons %>% 
  left_join(covid_incidence, c("KURZ" = "canton"))
```

To make our map look pretty, instead of using a numerical variable we divide the incidence into categories, so that it will be easier for the user to see in which category each canton is.

This a typical practice for choropleth maps, and it can be done in different ways.


```{r}
swiss_cantons <- swiss_cantons %>% 
  mutate(incidence_cat = case_when(
    incidence <= 50 ~ "0-50",
    incidence <= 100 ~ "51-100",
    incidence <= 150 ~ "101-150",
    incidence <= 300 ~ "251-300"
  )) %>% 
  mutate(incidence_cat = factor(incidence_cat, levels = c("0-50", "51-100","101-150","151-200","251-300")))
```

Now we can map the colour to the `incidence_cat` variable and create the first choropleth map.

```{r}
ggplot(swiss_cantons) +
  geom_sf(aes(fill = incidence_cat), size = 0.3) +
  scale_fill_carto_d(palette = "BrwnYl") +
  geom_sf(data = swiss_lakes, fill = "#d1eeea", color = "#d1eeea")+
  theme_void() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") 
```

And we have our first choropleth map, built only using `{sf}` and `{ggplot2}`.
Let add some finishing touches: we are not happy with how the legend looks like and we can change it using `guide_legend()`.

```{r}

ggplot(swiss_cantons) +
  geom_sf(aes(fill = incidence_cat), size = 0.3) +
  scale_fill_carto_d(palette = "BrwnYl",
                     guide = guide_legend(direction = "horizontal",
            keyheight = unit(2, units = "mm"),
            keywidth = unit(70 / 5, units = "mm"),
            title.position = 'top',
            title.hjust = 0.5,
            label.hjust = 0.5,
            nrow = 1,
            byrow = T,
            label.position = "bottom")) +
  geom_sf(data = swiss_lakes, fill = "#d1eeea", color = "#d1eeea")+
  geom_sf(data = swiss_border, fill= NA)+
  theme_void() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") 

```

Now we can add a title, a subtitle and labels on top of each canton.
We will use `{ggrepel}` to make sure the labels will not overlap with each other, we will also use `{ggtext}` so we can use markdown syntax for our title and subtitle.


```{r}
ggplot(swiss_cantons) +
  geom_sf(aes(fill = incidence_cat), size = 0.3) +
  scale_fill_carto_d(palette = "BrwnYl",
                     guide = guide_legend(direction = "horizontal",
            keyheight = unit(2, units = "mm"),
            keywidth = unit(70 / 5, units = "mm"),
            title.position = 'top',
            title.hjust = 0.5,
            label.hjust = 0.5,
            nrow = 1,
            byrow = T,
            label.position = "bottom")) +
  geom_sf(data = swiss_lakes, fill = "#d1eeea", color = "#d1eeea")+
  ggrepel::geom_label_repel(
    data = swiss_cantons,
    aes(label = paste0(KURZ,":",round(incidence, digits = 0)), 
        geometry = geometry),
    stat = "sf_coordinates",
    min.segment.length = 0.2,
    colour = "#541f3f",
    size = 3,
    segment.alpha = 0.5
  ) +
  labs(title = "<b style='color:#541f3f'> COVID-19 Cases by Canton, last 14 days </b>",
       subtitle = "<span style='font-size:10pt'>Incidence per 100'000 habitants,by canton </span>",
       caption = "Source: OFSP | updated 12.10.2020") +
  theme_void() +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        plot.title = ggtext::element_markdown(),
        plot.subtitle = ggtext::element_markdown()) 

```

We have seen now how to create a choropleth map, step by step using `{ggplot2}` and we have seen also how to customize it to make it look exactly how we want it to be.

